Human capital mediates natural selection in contemporary humans

### 2022-05-04 why-natural-selection.Rmd
Final final tweaks to pictures.

options(digits = 2)
options(dplyr.summarise.inform = FALSE) # shut up about summarize

@@

my_hline <- geom_hline(yintercept = 0, colour = "grey20", linetype = "dotted")
my_vline <- geom_vline(xintercept = 0, colour = "grey20", linetype = "dotted")
pval_palette <- c("FALSE" = "grey65", "TRUE" = "#f3b372")

@@

  n_regs <- as.double(n_regs)

  dfr %>% mutate(
    !! score_col := pretty_names(!! score_col),
    !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
  ) %>% 
    rename(p = p.value) %>% 
    ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
               shape = {{fill_col}}, ...)) +
    my_vline + 
    confint_geom_segment +
    geom_point(size = 1.8, alpha = 0.9, aes(color = {{fill_col}})) +
    geom_point(size = 0.7, color = "white", aes(alpha = p < 0.05/{{n_regs}})) +
    my_fill_scale(aesthetics = "color", n = n_cats, direction = fill_direction) +
    scale_shape_manual(values = 1:n_cats + 14) +
    my_labs() +
    my_theme() +
    confint_scale +
    scale_alpha_discrete(range = c(1, 0), 
                           guide = guide_legend(
                             order = 0,
                             override.aes = list(
                               shape = c(1, 16),
                               color = "black",
                               alpha = 1,
                               size = 1.8
                             )
                         )) 
}

@@

pval_scale <- function (...) scale_color_manual(values = pval_palette, ...)

@@

my_fill_scale <- function (n, aesthetics = "fill", ...) {

@@

  scale <- if (n <= 2) {
    dots <- list(...)
    dots$direction <- NULL
    do.call(scale_colour_manual, c(list(values = fill_colors[1:n],
                                        aesthetics = aesthetics), 
                                   dots))
  } else {
    scale <- scale_color_brewer(aesthetics = aesthetics, ...)
    shades::lightness(scale, shades::scalefac(0.75))
  }

@@

res_wt_combined %>% 
      filter(Weights %in% c("Main", "None")) %>% 
      mutate(
        Weights  = recode(Weights, "Main" = "Weighted", "None" = "Unweighted"),
        term     = pretty_names(term),
        term     = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      standard_ggplot(fill_col = Weights, n_regs = n_regs)

@@

res_edu %>% 
      mutate(
        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18"),
        "Age left FTE" = add_n(`Age left FTE`, "age_fte_cat")
      ) %>% 
      standard_ggplot(n_regs = n_regs, fill_col = `Age left FTE`, 
                        fill_direction = -1)

@@

n_regs <- as.double(nrow(res_income))
res_income %>% 
      mutate(
        Income = factor(income_cat, 
          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
        ),
        Income = add_n(Income, "income_cat")
      ) %>% 
      standard_ggplot(n_regs = n_regs, fill_col = `Income`, 
                        fill_direction = -1)
```

@@

n_regs <- as.double(nrow(res_partners_joint))
res_partners_joint %>% 
      mutate(
        `N partners` = add_n(`N partners`, "n_partners_split", reverse = TRUE,
                               data = famhist %>%
                               mutate(n_partners_split = n_partners <= 3))
      ) %>% 
      standard_ggplot(fill_col = `N partners`, n_regs = n_regs, 
                        score_col = score_name)

@@

res_with_partner %>% 
      mutate(
        Household = ifelse(grepl("TRUE", term), "With partner", 
                      "Without partner"),
        Household = add_n(Household, "with_partner", reverse = TRUE)
      ) %>% 
      standard_ggplot(fill_col = Household, n_regs = n_regs, 
                        score_col = score_name)
```

@@

res_age_flb_cross %>% 
      mutate(
        `Age at first live birth` = gsub("age_flb_cat(.*):.*", "\\1", term),
        `Age at first live birth` = add_n(`Age at first live birth`, "age_flb_cat")
      ) %>% 
      standard_ggplot(score_col = score_name, 
                      fill_col = `Age at first live birth`, n_regs = n_regs,
                      fill_direction = -1)

@@

dfr %>% 
      mutate(
        PGS      = ifelse(abs(effect_size) > 0.02, PGS, ""),
        PGS      = sub("(^\\w*?)_.*", "\\1", PGS)
      ) %>% 
      filter(
        Controls == "None"
      ) %>% 
      ggplot(aes(effect_size, cor, colour = Correlation)) + 
        geom_point() +
        geom_text(aes(label = PGS), size = 3, colour = "black", 
                    check_overlap = TRUE, nudge_x = 0.002) +
        facet_wrap( ~ Correlation, scales = "free") + 
        my_vline + 
        my_hline +
        labs(x = "Selection effect", y = "Correlation") +
        coord_cartesian(clip = "off") +
        scale_colour_brewer(type = "qual", palette = 2) +
        theme(
          legend.position = "none",
          panel.spacing = unit(20, "pt")
        )

@@

res_mediation %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, prop_ind, order_abs(1)),
        `Indirect effect p < 0.05/23` = sig 
      ) %>% 
      ggplot(aes(y = term)) + 
        geom_col(aes(x = prop_ind, fill = `Indirect effect p < 0.05/23`), 
                   width = .5) + 
        geom_errorbar(aes(
                          xmin = prop_ind_conf_low, 
                          xmax = prop_ind_conf_high,
                          linetype = "Bootstrap 95% c.i.",
                        ),
                        width = 0.3,
                        color = "grey50"
                      ) +
        scale_fill_manual(values = pval_palette) +
        my_vline +
        scale_x_continuous(labels = scales::percent, n.breaks = 7) +
        coord_cartesian(xlim = c(-0.5, 1)) +
        labs(x = "Proportion mediated", y = "") +
        scale_linetype_manual(name = "", values = c("Bootstrap 95% c.i." = 1))

@@

income_EA3 %>% 
              ggplot(aes(`Household income`, `Children's mean EA3`, fill = name)) +
                geom_col(position = "dodge") +
                scale_fill_manual(values = c("navy", "grey60")) +
                theme(
                  legend.background = element_rect(fill = "white", colour = NA),
                  legend.title = element_blank(), 
                  legend.position = c(.25, .85)
                )
```

@@

```{r plot-sexes, fig.cap = "Selection effects by sex. Highlighted lines are significant differences at p < 0.05/33. Highlighted points are significantly different from 0 at p < 0.05/66."}

@@

res_sex %>% 
      standard_ggplot(fill_col = sex, n_regs = n_regs, score_col = term,
                        conf_int = FALSE) +
      geom_linerange(data = res_sex_wide, inherit.aes = FALSE, 
                     color = "steelblue4", alpha = 1,
              aes(
                y      = term,
                x      = NULL,
                xmin   = min_est + 0.0005, 
                xmax   = max_est - 0.0005,
                linetype = p < 0.05/{{n_comps}}
              )) +
      scale_linetype_manual(values = c("TRUE" = 1, "FALSE" = 2))

@@

n_regs <- as.double(nrow(res_partners))
res_partners %>% 
      standard_ggplot(fill_col = `N partners`, score_col = score_name, 
                        n_regs = n_regs) +
        facet_grid(cols = vars(sex))

@@

res_with_partner_sex %>% 
      mutate(
        Household = ifelse(grepl("TRUE", term), "With partner", 
                      "Without partner")
      ) %>% 
      standard_ggplot(fill_col = Household, score_col = score_name, 
                        n_regs = n_regs) +
        facet_grid(cols = vars(sex))

@@

res_townsend_parents %>% 
    mutate(
        n = prettyNum(n, big.mark = ","),
        `Townsend quintile` = paste0(quintile, " (N = ", n, ")")
      ) %>% 
  standard_ggplot(fill_col    = `Townsend quintile`,
                    n_regs    = n_regs, 
                    score_col = score_name, 
                    order_idx = 5
                  )

@@

res_townsend %>% 
    mutate(
        n = prettyNum(n, big.mark = ","),
        `Townsend quintile` = paste0(quintile, " (N = ", n, ")")
      ) %>% 
  standard_ggplot(fill_col    = `Townsend quintile`,
                    n_regs    = n_regs, 
                    score_col = score_name, 
                    order_idx = 5
                  )

@@

res_age_flb_mothers_cross %>% 
      mutate(
        `Age at first live birth` = gsub("moth_age_birth_cat(.*):.*", "\\1", term),
        `Age at first live birth` = add_n(`Age at first live birth`, 
                                            "moth_age_birth_cat", 
                                            dv   = "RLRS_parents",
                                            data = famhist_bo1
                                          )
      ) %>% 
      standard_ggplot(score_col = score_name, 
                      fill_col = `Age at first live birth`, n_regs = n_regs,
                      fill_direction = -1)

@@

res_age_flb_fathers_cross %>% 
      mutate(
        `Age at first live birth` = gsub("fath_age_birth_cat(.*):.*", "\\1", term),
        `Age at first live birth` = add_n(`Age at first live birth`, 
                                            "fath_age_birth_cat", 
                                            dv   = "RLRS_parents",
                                            data = famhist_bo1
                                          )
      ) %>% 
      standard_ggplot(score_col = score_name, 
                      fill_col = `Age at first live birth`, n_regs = n_regs,
                      fill_direction = -1)

@@

res_age_birth_parents %>% 
    filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
    mutate(
      Control = ifelse(control == "fath_age_birth", 
                  "Father's age at birth", 
                  "Mother's age at birth"
                )
    ) %>% 
    standard_ggplot(fill_col = Control, n_regs = n_regs, order_idx = 2)

@@

```{r plot-age-flb-dv, fig.cap = "Effects of polygenic scores on age at first live birth."}

@@

```{r plot-age-birth-parents-dv, fig.cap = "Effects of polygenic scores on parents' age at respondent's birth, eldest siblings. Weights multiplied by 1/number of siblings."}

@@

res_age_birth_parents_dv %>% 
      mutate(
        "Dependent variable" = dplyr::recode(dep.var, 
          "fath_age_birth" = "Father's AFLB", 
          "moth_age_birth" = "Mother's AFLB"
        )
      ) %>% 
      standard_ggplot(fill_col = `Dependent variable`, n_regs = n_regs)

@@

res_mediation %<>% mutate(
                     # 0.025 not 0.05 because two-sided
                     sig  = abs(statistic_ind) > qnorm(1 - 0.025/33),
                     star = ifelse(sig, "*", "")
                   )

@@

footnote <- sprintf("* p < 0.05/33. Analysis run on %s PGS which correlated 
                    significantly with fertility.", nrow(res_mediation))

@@

Table \@ref(tab:tbl-res-mediation) shows results. For `r n_pos_sig` out of
`r nrow(res_mediation)` scores, the indirect effect on fertility via human 
capital is significantly different from 0 at $p = 0.05/33$ and has the
same sign as the total effect. We also calculate
the proportion of the total effect that is mediated via the indirect effect,
along with uncorrected 95% confidence intervals (100 bootstraps).

@@

res_all %>% 
      filter(dep.var == "RLRS") %>% 
      standard_ggplot(fill_col = PGS, n_regs = n_regs)

res_all %>% 
      filter(dep.var == "RLRS_parents") %>% 
      standard_ggplot(fill_col = PGS, n_regs = n_regs)

>>>>

options(digits = 2)
options(dplyr.summarise.inform = FALSE)

@@

my_hline <- geom_hline(yintercept = 0, colour = "grey20", linetype = "dotted")
my_vline <- geom_vline(xintercept = 0, colour = "grey20", linetype = "dotted")

@@

  n_regs <- as.double(n_regs)
  shape_values <- c(16, 15, 17, 18, 20)[1:n_cats]
  if (fill_direction == -1) shape_values <- rev(shape_values)
  dfr %>% mutate(
    !! score_col := pretty_names(!! score_col),
    !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
  ) %>% 
    rename(p = p.value) %>% 
    ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
               shape = {{fill_col}}, ...)) +
    my_vline + 
    confint_geom_segment +
    geom_point(size = 1.8, alpha = 0.9, aes(color = {{fill_col}})) +
    geom_point(size = 0.7, color = "white", aes(alpha = p < 0.05/{{n_regs}})) +
    my_fill_scale(aesthetics = "color", n = n_cats, direction = fill_direction) +
    scale_shape_manual(values = shape_values) +
    my_labs() +
    my_theme() +
    confint_scale +
    scale_alpha_discrete(range = c(1, 0), 
                           guide = guide_legend(
                             order = 0,
                             override.aes = list(
                               shape = c(1, 16),
                               color = "black",
                               alpha = 1,
                               size = 1.8
                             )
                         )) 
}

@@

@@

my_fill_scale <- function (n, aesthetics = "fill", direction = 1, ...) {

@@

  scale <- if (n <= 2) {
    dots <- list(...)
    dots$direction <- NULL
    do.call(scale_colour_manual, c(list(values = fill_colors[1:n],
                                        aesthetics = aesthetics), 
                                   dots))
  } else {
    # scale <- scale_color_viridis_d(aesthetics = aesthetics, option = "D",
    #                                  direction = direction, end = 0.6, ...)
    if (direction != -1) fill_colors <- rev(fill_colors)
    scale_color_manual(aesthetics = aesthetics, 
                         values = shades::gradient(fill_colors,
                                                     steps = n), 
                         ...) 
  }

@@

res_wt_combined %>% 
      filter(Weights %in% c("Main", "None")) %>% 
      mutate(
        Weights  = recode(Weights, "Main" = "Weighted", "None" = "Unweighted"),
        term     = pretty_names(term),
        term     = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      standard_ggplot(fill_col = Weights, n_regs = n_regs, fill_direction = -1)

@@

res_edu %>% 
      mutate(
        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18"),
        "Age left FTE" = add_n(`Age left FTE`, "age_fte_cat")
      ) %>% 
      standard_ggplot(n_regs = n_regs, fill_col = `Age left FTE`)

@@

n_regs <- as.double(nrow(res_income))
res_income %>% 
      mutate(
        Income = factor(income_cat, 
          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
        ),
        Income = add_n(Income, "income_cat")
      ) %>% 
      standard_ggplot(n_regs = n_regs, fill_col = `Income`)
```

@@

n_regs <- as.double(nrow(res_partners_joint))
res_partners_joint %>% 
      mutate(
        `N partners` = add_n(`N partners`, "n_partners_split", reverse = TRUE,
                               data = famhist %>%
                               mutate(n_partners_split = n_partners <= 3))
      ) %>% 
      standard_ggplot(fill_col = `N partners`, n_regs = n_regs, 
                        score_col = score_name, fill_direction = -1)

@@

res_with_partner %>% 
      mutate(
        Household = ifelse(grepl("TRUE", term), "With partner", 
                      "Without partner"),
        Household = add_n(Household, "with_partner", reverse = TRUE)
      ) %>% 
      standard_ggplot(fill_col = Household, n_regs = n_regs, 
                        score_col = score_name, fill_direction = -1)
```

@@

res_age_flb_cross %>% 
      mutate(
        `Age at first live birth` = gsub("age_flb_cat(.*):.*", "\\1", term),
        `Age at first live birth` = add_n(`Age at first live birth`, "age_flb_cat")
      ) %>% 
      standard_ggplot(score_col = score_name, 
                      fill_col = `Age at first live birth`, n_regs = n_regs)

@@

dfr %>% 
      mutate(
        PGS      = ifelse(abs(effect_size) > 0.027, PGS, ""),
        PGS      = sub("(^\\w*?)_.*", "\\1", PGS)
      ) %>% 
      filter(
        Controls == "None"
      ) %>% 
      ggplot(aes(effect_size, cor)) + 
        geom_point(color = "steelblue4") +
        geom_text(aes(label = PGS), size = 3, colour = "black", 
                    check_overlap = TRUE, nudge_x = 0.002) +
        facet_wrap( ~ Correlation, scales = "free") + 
        my_vline + 
        my_hline +
        labs(x = "Selection effect", y = "Correlation") +
        coord_cartesian(clip = "off") +
        theme(
          legend.position = "none",
          panel.spacing = unit(20, "pt")
        )

@@

res_mediation %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, prop_ind, order_abs(1)),
        `Indirect effect p < 0.05/23` = sig 
      ) %>% 
      ggplot(aes(y = term)) + 
        geom_col(aes(x = prop_ind, fill = `Indirect effect p < 0.05/23`), 
                   width = .5) + 
        geom_errorbar(aes(
                          xmin = prop_ind_conf_low, 
                          xmax = prop_ind_conf_high,
                          linetype = "Bootstrap 95% c.i.",
                        ),
                        width = 0.3,
                        color = "grey50"
                      ) +
        scale_fill_manual(values = c("FALSE" = "grey65", "TRUE" = "orange2")) +
        my_vline +
        scale_x_continuous(labels = scales::percent, n.breaks = 7) +
        coord_cartesian(xlim = c(-0.5, 1)) +
        labs(x = "Proportion mediated", y = "") +
        scale_linetype_manual(name = "", values = c("Bootstrap 95% c.i." = 1))

@@

income_EA3 %>% 
              ggplot(aes(`Household income`, `Children's mean EA3`, fill = name)) +
                geom_col(position = "dodge") +
                scale_fill_manual(values = c("darkred", "grey65")) +
                theme(
                  legend.background = element_rect(fill = "white", colour = NA),
                  legend.title = element_blank(), 
                  legend.position = c(.25, .85)
                )
```

@@

```{r plot-sexes, fig.cap = "Selection effects by sex. Solid lines are significant differences at p < 0.05/33. Solid points are significantly different from 0 at p < 0.05/66."}

@@

res_sex %>% 
      standard_ggplot(fill_col = sex, n_regs = n_regs, score_col = term,
                        conf_int = FALSE) +
      geom_linerange(data = res_sex_wide, inherit.aes = FALSE, 
                     color = "steelblue4", alpha = 1,
              aes(
                y      = term,
                x      = NULL,
                xmin   = min_est + 0.0005, 
                xmax   = max_est - 0.0005,
                linetype = p < 0.05/{{n_comps}}
              )) +
      scale_linetype_manual(values = c("TRUE" = 1, "FALSE" = 3))

@@

n_regs <- as.double(nrow(res_partners))
res_partners %>% 
      standard_ggplot(fill_col = `N partners`, score_col = score_name, 
                        n_regs = n_regs, fill_direction = -1) +
        facet_grid(cols = vars(sex))

@@

res_with_partner_sex %>% 
      mutate(
        Household = ifelse(grepl("TRUE", term), "With partner", 
                      "Without partner")
      ) %>% 
      standard_ggplot(fill_col = Household, score_col = score_name, 
                        n_regs = n_regs, fill_direction = -1) +
        facet_grid(cols = vars(sex))

@@

res_townsend_parents %>% 
    mutate(
        n = prettyNum(n, big.mark = ","),
        `Townsend quintile` = paste0(quintile, " (N = ", n, ")")
      ) %>% 
  standard_ggplot(fill_col    = `Townsend quintile`,
                    n_regs    = n_regs, 
                    score_col = score_name, 
                    order_idx = 5,
                    fill_direction = -1
                  )

@@

res_townsend %>% 
    mutate(
        n = prettyNum(n, big.mark = ","),
        `Townsend quintile` = paste0(quintile, " (N = ", n, ")")
      ) %>% 
  standard_ggplot(fill_col    = `Townsend quintile`,
                    n_regs    = n_regs, 
                    score_col = score_name, 
                    order_idx = 5,
                    fill_direction = -1
                  )

@@

res_age_flb_mothers_cross %>% 
      mutate(
        `Age at first live birth` = gsub("moth_age_birth_cat(.*):.*", "\\1", term),
        `Age at first live birth` = add_n(`Age at first live birth`, 
                                            "moth_age_birth_cat", 
                                            dv   = "RLRS_parents",
                                            data = famhist_bo1
                                          )
      ) %>% 
      standard_ggplot(score_col = score_name, 
                      fill_col = `Age at first live birth`, n_regs = n_regs)

@@

res_age_flb_fathers_cross %>% 
      mutate(
        `Age at first live birth` = gsub("fath_age_birth_cat(.*):.*", "\\1", term),
        `Age at first live birth` = add_n(`Age at first live birth`, 
                                            "fath_age_birth_cat", 
                                            dv   = "RLRS_parents",
                                            data = famhist_bo1
                                          )
      ) %>% 
      standard_ggplot(score_col = score_name, 
                      fill_col = `Age at first live birth`, n_regs = n_regs)

@@

res_age_birth_parents %>% 
    filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
    mutate(
      Control = ifelse(control == "fath_age_birth", 
                  "Father's age at birth", 
                  "Mother's age at birth"
                )
    ) %>% 
    standard_ggplot(fill_col = Control, n_regs = n_regs, order_idx = 2,
                      fill_direction = -1)

@@

```{r plot-age-flb-dv, fig.cap = "Effects of polygenic scores on age at first live birth.", fig.align = "center"}

drake::loadd(res_age_flb_dv)

@@

```{r plot-age-birth-parents-dv, fig.cap = "Effects of polygenic scores on parents' age at respondent's birth, eldest siblings. Weights multiplied by 1/number of siblings.", fig.align = "center"}

drake::loadd(res_age_birth_parents_dv)

@@

res_age_birth_parents_dv %>% 
      mutate(
        "Dependent variable" = dplyr::recode(dep.var, 
          "fath_age_birth" = "Father's AFLB", 
          "moth_age_birth" = "Mother's AFLB"
        )
      ) %>% 
      standard_ggplot(fill_col = `Dependent variable`, n_regs = n_regs, 
                        fill_direction = -1)

@@

n_regs <- nrow(res_mediation)
res_mediation %<>% mutate(
                     # 0.025 not 0.05 because two-sided
                     sig  = abs(statistic_ind) > qnorm(1 - 0.025/n_regs),
                     star = ifelse(sig, "*", "")
                   )

@@

footnote <- sprintf("* p < 0.05/%s. Analysis run on %s PGS which correlated significantly with fertility.", n_regs, n_regs)

@@

Table \@ref(tab:tbl-res-mediation) shows results. For `r n_pos_sig` out of 
`r n_regs` scores, the indirect effect on fertility via human capital is
significantly different from 0 at $p$ = 0.05/`r n_regs` and has the same sign
as the total effect. We also calculate the proportion of the total effect that
is mediated via the indirect effect, along with uncorrected 95% confidence
intervals (100 bootstraps).

@@

res_all %>% 
      filter(dep.var == "RLRS") %>% 
      standard_ggplot(fill_col = PGS, n_regs = n_regs, fill_direction = -1)

res_all %>% 
      filter(dep.var == "RLRS_parents") %>% 
      standard_ggplot(fill_col = PGS, n_regs = n_regs, fill_direction = -1)
